Combining checkpointing and data compression\\ for high frequency seismic imaging

### 2018-10-01 paper/paper.tex
Add performance model to paper

\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

@@

p-adaptive compression from Boehm et al. 
\section{Results}
\subsection{Error analysis}
Discuss the error incurred due to lossy compression. This could be the
angular deviation in the gradient from Boehm et al. as a factor of the
precision parameter, for about 3 different compression schemes  What if a minor
deviation is at a point where you're pointing directly at a
constraint? or a corner point between two constraints? The line search
may mistakenly terminate early. This should be used to derive a bound
for compression ratio based on desired precision.

>>>>

\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{cuted}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

One possible way of using compression to speed up full-waveform inversion or any adjoint-based
optimization is by simply compressing the field data before storing it and decompressing it when it
is required again. If the memory available is sufficient to store the complete forward data after
compression, the use of compression effectively replaces the use of revolve-based checkpointing.
This scenario has been discussed in different studies before - some of which we cover in the next
section. This paper focusses on a broader region, of which this scenario is just one part. In particular,
we discuss the scenarios where the forward field can not fit in memory even after compression, where
compression and Revolve would be used in conjunction.

Although most of the compression we discuss here introduces some error which may or may not affect
the final result of the computation, we try to avoid discussing the error in this paper and focus on the 
performance impact of compression instead.

@@

p-adaptive compression from Boehm et al.

\section{Compression in adjoint-based optimization}
We work with the assumption that the computation of a single forward time step takes the same wall time
as the computation of a single reverse time step - calling this $\mathbf{C}$. If the size of a single 
timestep in memory is $\mathbf{S}$ and the simulation involves $\mathbf{N}$ timesteps, the minimum
wall time required to run a single forward-adjoint evaluation is given by:
\begin{equation}
T_N = 2 \times \mathbf{C} \times \mathbf{N}
\end{equation}
However, this would require $\mathbf{S} \times \mathbf{N}$ memory. On a system where enough memory
is available, this would obviously be the optimal strategy to compute the adjoint solution.

When the memory available on a system is less than $\mathbf{S} \times \mathbf{N}$, Revolve provides
an optimal strategy to solve for the adjoint field by storing a subset of the $\mathbf{N}$ total checkpoints
and recompute the remaining ones. The overhead introduced by this method can be broken down into
the recomputation overhead $\mathbf{O}_R$ and the storage overhead $\mathbf{O}_S$. The recomputation
overhead is largely fixed and given by the following relationship (TODO cite revolve):
\begin{strip}
\begin{equation}
\mathbf{O}_R(N, M) = \begin{cases}
      N(N-1) /2, & \text{if}\ M=1 \\
      \min\limits_{1<=\widetilde{N}<=N} \{\widetilde{N} + \mathbf{O}_R(\widetilde{N}, M) + \mathbf{O}_R(N-\widetilde{N}, M-1)\}, & \text{if}\ M>1
    \end{cases}
    \label{eqn:recompute}
\end{equation}
\end{strip}
where M is the number of checkpoints that can be stored in memory. Note that for $M >=N$, $\mathbf{O}_R$
would be zero. For $M < N$, $\mathbf{O}_R$ goes up rather quickly as M is reduced in relation to N.

In a perfect implementation, the storage overhead $\mathbf{O}_S$ might be zero, since the computation could
be done "in-place", but in practice, checkpoints are generally stored in a separate section of memory and they
need to be transferred to a "computational" section of the memory where the compute is performed, and then
the results copied back to the checkpointing memory. This copying is a common feature of checkpointing
implementations, and might pose a non-trivial overhead in scenarios where the computation is heavily bandwidth-bound. 
This storage overhead is given by:
\begin{equation}
\mathbf{O}_SR = \mathbf{N}_W \times \frac{\mathbf{S}}{\mathbf{B}} + \mathbf{N}_R \times \frac{\mathbf{S}}{\mathbf{B}}
\label{eqn:storage}
\end{equation}
where $\mathbf{N}_W$ is the total number of times Revolve writes checkpoints for a single run, $ \mathbf{N}_R$ 
is the number of times checkpoints are read, and $\mathbf{B}$ is the memory bandwidth of the target system.

The total time to solution in this scenario becomes:
\begin{equation}
T_R = 2 \times \mathbf{C} \times \mathbf{N} + \mathbf{O}_R(N, M) + \mathbf{O}_S
\end{equation}

Adding compression to the mix will increase the number of checkpoints available ($M$ in equation \ref{eqn:recompute}),
hence reducing $\mathbf{O}_R$ while adding overheads related to compression and decompression in $\mathbf{O}_S$,
as well as some error introduced by the compression. Assuming that the error remains within the allowed tolerance,
we hypothesize that, at least for some circumstances, the reduction in $\mathbf{O}_R$ should offset the increase in 
$\mathbf{O}_S$, i.e. lead to an overall decrease in the total time to solution ($T$). To build a performance model that
enables us to find the regions where compression pays off, we assume that the compression algorithm behaves uniformly
across the different time steps of the simulation, i.e. that we get the same compression ratio, compression time and 
decompression time, no matter which of the $N$ possible checkpoints we try to compress/decompress. The storage overhead
now becomes:
\begin{equation}
\mathbf{O}_SR = \mathbf{N}_W \times (\frac{\mathbf{S}}{\mathbf{F}\mathbf{B}} + t_c) + \mathbf{N}_R \times (\frac{\mathbf{S}}{\mathbf{F}\mathbf{B}} + t_d)
\end{equation}
where $\mathbf{F}$ is the compression ratio (i.e. the ratio between the uncompressed and compressed checkpoint), and $t_c$
and $t_d$ are compression and decompression times, respectively. At the same time, the recomputation overhead goes down
because $\mathbf{F}$ times more checkpoints are now available.

TODO add figure here that shows memory on x-axis and speedup on y axis. Split into 3 regions. Region 1, 
where Revolve+Compression makes sense, Region 2, where compression would probably not bring in enough benefit, 
and region 3 where compression would enable the computation to be done without revolve. 
\section{Performance model}
We can predict the performance
\section{Results}
\subsection{Error analysis}
Discuss the error incurred due to lossy compression. This could be the
angular deviation in the gradient from Boehm et al. as a factor of the
precision parameter, for about 3 different compression schemes  What if a minor
deviation is at a point where you're pointing directly at a
constraint? or a corner point between two constraints? The line search
may mistakenly terminate early. This should be used to derive a bound
for compression ratio based on desired precision.
